Parties, Models, Mobsters\newline {\fontsize{11}{15}\selectfont A New Implementation of Model-Based Recursive Partitioning in R}

### 2014-03-11 pkg/partykit/inst/NUS-2014/modelparty.Rnw
improved theory section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@467 255f4249-2263-4417-b1ea-8498779e27a7

\textbf{Examples:} Recursive partitioning, decision trees.
Class of flexible methods for classification and regression.

\textbf{Formally}: Modeling of dependent variable $Y$ by ``learning''
a recursive partition w.r.t\ explanatory variables $Z_1, \dots, Z_l$.

\textbf{Key features}:

\begin{enumerate}
  \item Predictive power in nonlinear regression relationships.
  \item Interpretability (enhanced by visualization), i.e., no ``black box'' methods.
\end{enumerate}

\textbf{Examples}: CART and C4.5 in statistical and machine learning, respectively.

@@

\textbf{Typically:} Simple models for univeriate $Y$, e.g., mean.

@@

\textbf{Idea:} More complex models for multivariate $Y$, e.g., multivariate
normal model, regression models, etc.

@@

\textbf{Base algorithm}:

\begin{enumerate}
  \item Fit model for $Y$.
  \item Assess association of $Y$ and each $Z_j$.
  \item Split sample along the $Z_{j^*}$ with strongest association:
        Choose breakpoint with highest improvement of the model fit.
  \item Repeat steps 1--3 recursively in the subsamples
        until some stopping criterion is met.
\end{enumerate}

%% \medskip
%% 
%% \textbf{Generally:} Tree algorithms differ w.r.t.\ choice of model (1),
%% association measure~(2), split strategy (3) and stopping criterion (4).

\textbf{Here:} Segmentation~(3) of parametric models~(1) with 
additive objective function using parameter instability tests~(2)
and associated statistical significance~(4).

@@

\textbf{Models:} $\mathcal{M}(Y, \theta)$ with (potentially) multivariate
observations $Y \in \mathcal{Y}$ and $k$-dimensional parameter vector
$\theta \in \Theta$.

@@

\textbf{Parameter estimation:} $\widehat \theta$ by optimization of
objective function $\Psi(Y, \theta)$ for $n$ observations $Y_i$ ($i = 1, \dots, n$):

\[ \widehat \theta \quad = \quad \argmin_{\theta \in \Theta}
   \sum_{i = 1}^n \Psi(Y_i, \theta). \]

@@

\textbf{Estimating function:} $\widehat{\theta}$ can also be defined in terms of
  \[ \sum_{i = 1}^n \psi(Y_i, \widehat{\theta}) = 0, \]
where $\psi(Y, \theta) = {\partial \Psi(Y, \theta)}/{\partial \theta}$.

@@

\textbf{Idea:}
In many situations, a single global model 
$\mathcal{M}(Y, \theta)$ that fits \textbf{all} $n$~observations cannot be found.
But it might be possible to find a partition w.r.t.\ the variables
$Z = (Z_1, \dots, Z_l)$ so that a well-fitting model can be found locally
in each cell of the partition.

@@

\textbf{Tool:} Assess parameter instability w.r.t\ to partitioning variables
$Z_j \in \mathcal{Z}_j \; (j = 1, \dots, l)$.

@@

Generalized M-fluctuation tests
capture instabilities in $\widehat \theta$ for an ordering w.r.t\ $Z_j$.

@@

\textbf{Basis:} Empirical fluctuation process of cumulative deviations
w.r.t.\ to an ordering $\sigma(Z_{ij})$.

\[
W_j(t, \widehat \theta)
  \quad = \quad {\widehat B}^{-1/2} n^{-1/2}
                \sum_{i = 1}^{\lfloor nt \rfloor}
		\psi(Y_{\sigma(Z_{ij})}, \widehat \theta)
\qquad (0 \le t \le 1)
\]

@@

## data, model, and residuals
plot(0:81, x, xlab = expression(Z[j]), ylab = "Y", ylim = c(-2, 3),
  type = "l", xlim = c(1, 80), axes=FALSE)
axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
axis(2,at=seq(-2,3,1),labels=seq(200,1200,200),cex.axis=1.2, cex.lab=1.2)
box()
abline(h = mean(x), lty = 3)
for(i in c(5, 15, 25, 35)) {
  segments(x0 = i, x1 = i, y0 = x[i+1], y1 = mean(x), lty = 2)
  i <- i + 40
  segments(x0 = i, x1 = i, y0 = mean(x), y1 = x[i+1], lty = 2)
}

## CUSUM graph
x_efp <- c(0, cumsum(x - mean(x)))/sqrt(length(x) * var(x))
plot(-1:81, x_efp, xlab = expression(Z[j]), ylab = "Fluctuation Process", 
  type = "l", xlim = c(1, 80), ylim = c(-5, 5), axes=FALSE)
axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
axis(2, cex.axis=1.2, cex.lab=1.2)
box()
abline(h = 0, lty = 3)
for(i in seq(5, 75, by = 10)) segments(x0 = i, x1 = i, y0 = 0, y1 = x_efp[i+2], lty = 2)
@

@@

\textbf{Special cases:} Class of test encompasses many well-known
tests for different classes of models.
Certain functionals~$\lambda$ are particularly intuitive for numeric and
categorical $Z_j$, respectively.

@@

\textbf{Advantage:} Model $\mathcal{M}(Y, \widehat \theta)$ just has to be
estimated once. Empirical estimating functions $\psi(Y_i, \widehat \theta)$
just have to be re-ordered and aggregated for each $Z_j$.

@@

\[
\lambda_{\mbox{sup}\mathit{LM}} (W_j) \quad = \quad
\max_{i = \ui, \dots, \oi}  \, \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1}
\left| \left| W_j \left( \frac{i}{n} \right) \right| \right|^2_2.
\]

@@

@@

\[ \lambda_{\chi^2} (W_j) \quad = \quad
\sum_{c = 1}^C 
\frac{n}{|I_c|}  \left| \left| \Delta_{I_c} W_j \left( \frac{i}{n} \right) \right| \right|^2_2
\]

@@

\textbf{Splitting ordinal variables:} FIXME

like categorical

like numeric but invariant to reordering within ties

@@

\textbf{Goal:} Split model into $b = 1, \dots, B$ segments along the
partitioning variable $Z_j$ associated with the highest parameter instability.
Local optimization of

\[ \sum_b \sum_{i \in I_b} \Psi(Y_i, \theta_b). \]

@@

\textbf{Here:} Binary partitioning.

@@

\textbf{Pruning:} Avoid overfitting.

@@

\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when
there is no significant parameter instability based on
Bonferroni-corrected $p$~values of the fluctuation tests.

@@

\textbf{Post-pruning:} Grow large tree and prune splits that do not
improve the model fit based on information criteria (e.g., AIC or BIC).

>>>>

\textbf{Classical example}: Trees, i.e., modeling of dependent variable $y$ by ``learning''
a recursive partition w.r.t\ explanatory variables $z_1, \dots, z_l$.

@@

\textbf{Key features}:
\begin{enumerate}
  \item Predictive power in nonlinear regression relationships.
  \item Interpretability (enhanced by visualization), i.e., no ``black box'' methods.
\end{enumerate}

@@

\textbf{Typically:} Simple models for univeriate $y$, e.g., mean.

\textbf{Idea:} More complex models for more complex $y$, e.g.,
regression models, multivariate normal model, item responses, etc.

@@

\textbf{Model-based algorithm}:

\begin{enumerate}
  \item Fit the parametric model in the current subsample.
  \item Assess the stability of the parameters across each partitioning variable $z_j$.
  \item Split sample along the $z_{j^*}$ with strongest instability:
        Choose breakpoint with highest improvement of the model fit.
  \item Repeat steps 1--3 recursively in the subsamples
        until some stopping criterion is met.
\end{enumerate}

@@

\textbf{Models:} $\mathcal{M}(y, x, \theta)$ with (potentially multivariate)
observations $y$, optionally regressors $x$, and $k$-dimensional parameter vector
$\theta \in \Theta$.

@@

\textbf{Parameter estimation:} $\widehat \theta$ by optimization of
objective function $\Psi(y, x, \theta)$ for $n$ observations $y_i$ ($i = 1, \dots, n$):

\[ \widehat \theta \quad = \quad \argmin_{\theta \in \Theta}
   \sum_{i = 1}^n \Psi(y_i, x_i, \theta). \]

@@

\textbf{Estimating function:} $\widehat{\theta}$ can also be defined in terms of
  \[ \sum_{i = 1}^n \psi(y_i, x_i, \widehat{\theta}) = 0, \]
where $\psi(y, x, \theta) = {\partial \Psi(y, x, \theta)}/{\partial \theta}$.

@@

\textbf{Idea:}
In many situations, a single global model 
$\mathcal{M}(y, x, \theta)$ that fits \textbf{all} $n$~observations cannot be found.
But it might be possible to find a partition w.r.t.\ the variables
$z_1, \dots, z_l$ so that a well-fitting model can be found locally
in each cell of the partition.

@@

\textbf{Tools:}
\begin{itemize}
  \item Assess parameter instability w.r.t\ to partitioning variables
        $z_j \; (j = 1, \dots, l)$.
  \item A general measure of deviation from the model is the estimating function $\psi(y, x, \theta)$.
\end{itemize}

@@

Generalized M-fluctuation tests
capture instabilities in $\widehat \theta$ for an ordering w.r.t\ $z_j$.

@@

\textbf{Basis:} Empirical fluctuation process of cumulative deviations
w.r.t.\ to an ordering $\sigma(z_{ij})$.

\[
W_j(t, \widehat \theta)
  \quad = \quad {\widehat B}^{-1/2} n^{-1/2}
                \sum_{i = 1}^{\lfloor nt \rfloor}
		\psi(y_{\sigma(z_{ij})}, x_{\sigma(z_{ij})}, \widehat \theta)
\qquad (0 \le t \le 1)
\]

@@

## data, model, and residuals
plot(0:81, x, xlab = expression(z[j]), ylab = "y", ylim = c(-2, 3),
  type = "l", xlim = c(1, 80), axes=FALSE)
axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
axis(2,at=seq(-2,3,1),labels=seq(200,1200,200),cex.axis=1.2, cex.lab=1.2)
box()
abline(h = mean(x), lty = 3)
for(i in c(5, 15, 25, 35)) {
  segments(x0 = i, x1 = i, y0 = x[i+1], y1 = mean(x), lty = 2)
  i <- i + 40
  segments(x0 = i, x1 = i, y0 = mean(x), y1 = x[i+1], lty = 2)
}

## CUSUM graph
x_efp <- c(0, cumsum(x - mean(x)))/sqrt(length(x) * var(x))
plot(-1:81, x_efp, xlab = expression(z[j]), ylab = "Fluctuation Process", 
  type = "l", xlim = c(1, 80), ylim = c(-5, 5), axes=FALSE)
axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
axis(2, cex.axis=1.2, cex.lab=1.2)
box()
abline(h = 0, lty = 3)
for(i in seq(5, 75, by = 10)) segments(x0 = i, x1 = i, y0 = 0, y1 = x_efp[i+2], lty = 2)
@

@@

\textbf{Special cases:} Class of test encompasses many well-known
tests for different classes of models.
Certain functionals~$\lambda$ are particularly intuitive for numeric and
categorical $z_j$, respectively.

@@

\textbf{Advantage:} Model $\mathcal{M}(y, x, \widehat \theta)$ just has to be
estimated once. Empirical estimating functions $\psi(y_i, x_i, \widehat \theta)$
just have to be re-ordered and aggregated for each $z_j$.

@@

\[
\lambda_{\mathit{supLM}} (W_j) \quad = \quad
\max_{i = \ui, \dots, \oi}  \, \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1}
\left| \left| W_j \left( \frac{i}{n} \right) \right| \right|^2_2.
\]

@@

\textbf{Potential alternatives:} Many other parameter instability tests from the same
class of tests, e.g., a Cram\'{e}r-von Mises test (or Nyblom-Hansen test),
MOSUM tests, etc.

@@

\[ \lambda_{\chi^2} (W_j) \quad = \quad
\sum_{c = 1}^C 
\frac{n}{|I_c|}  \left| \left| \Delta_{I_c} W_j \left( \frac{i}{n} \right) \right| \right|^2_2.
\]

@@

\textbf{Splitting ordinal variables:} Several strategies conceivable.
Assess instability either as for categorical variables (if $C$ is low),
or as for numeric variables (if $C$ is high), or via a specialized test.
\begin{eqnarray*}
  \lambda_\mathit{maxLMo}(W_j) & = & \max_{i \in \{i_1, \dots, i_{C-1} \}} ~
    \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1}
    \left| \left| W_j \left( \frac{i}{n} \right) \right| \right|^2_2,\\
  \lambda_\mathit{WDMo}(W_j) & = & \max_{i \in \{i_1, \dots, i_{C-1} \}} ~
    \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1/2}      
    \left| \left| W_j\left( \frac{i}{n} \right) \right| \right|_{\infty}.
\end{eqnarray*}

\textbf{Interpretation:} Assess only the possible splitpoints $i_1, \dots, i_{C-1}$,
based on $L_2$ or $L_\infty$ norm.

\textbf{Limiting distribution:} Maximum from selected points in a squared
Bessel process or multivariate normal distribution, respectively.

@@

\textbf{Goal:} Split model into $b = 1, \dots, B$ segments along the
partitioning variable $z_j$ associated with the highest parameter instability.
Local optimization of

\[ \sum_b \sum_{i \in I_b} \Psi(y_i, x_i, \theta_b). \]

@@

\textbf{Here:} Binary partitioning. Optionally, $B = C$ can be chosen (without search)
for categorical variables.

@@

\textbf{Goal:} Avoid overfitting.

@@

\textbf{Pre-pruning:}
\begin{itemize}
  \item Internal stopping criterium.
  \item Stop splitting when there is no significant parameter instability.
  \item Based on Bonferroni-corrected $p$~values of the fluctuation tests.
\end{itemize}

@@

\textbf{Post-pruning:}
\begin{itemize}
  \item Grow large tree (e.g., with high significance level).
  \item Prune splits that do not improve the model fit
        based on information criteria (e.g., AIC or BIC).
\end{itemize}

\textbf{Hyperparameters:} Significance level and information criterion penalty
can be chosen manually (or possibly through cross-validation etc.).
